---
title: "Mini-Project3_YangJ_XieX"
author: "Jackie Yang, Xinyu Xie"
date: "2025-04-28"
output: html_document
---

## Front Matter

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(MASS)
library(glmnet)
library(rpart)
library(rattle)

CODGames2_mp <- read_excel("C:/ESSENTIALS/School OS/Junior Spring 2025/STAT 380/Datasets/CODGames2_mp.xlsx")
```

## Task 1 - Remove Partial

```{r}
CODGames2_mp <- CODGames2_mp %>%
  filter(FullPartial != "Partial")
CODGames2_mp

ggplot(CODGames2_mp, aes(x = XPType, y = TotalXP)) + 
  geom_boxplot(fill = "lightblue") + 
  labs(x = "Experience Type", y = "Total Experience", title = "Total Experience by Experience Type")

CODGames2_mp %>%
  group_by(XPType) %>%
  summarize(
    Mean = mean(TotalXP, na.rm = TRUE),
    Median = median(TotalXP, na.rm = TRUE),
    SD = sd(TotalXP, na.rm = TRUE),
    Min = min(TotalXP, na.rm = TRUE),
    Max = max(TotalXP, na.rm = TRUE),
    Count = n()
  )
```

Based on the summary statistics, the games where the player had Double XP + 10% as the experience type generally had more total experience than when the player had only the 10% Boost. Also, looking at the boxplots, the distribution of the experience type DOuble XP + 10% shows a higher spread than the experience type 10% Boost. Additionally, there seems to be more outliers.

## Task 2 Building Model

```{r}
CODGames2_mp_HC_TDM <- CODGames2_mp %>%
  filter(GameType == "HC - TDM")

CODGames2_mp_HC_TDM <- CODGames2_mp_HC_TDM %>%
  separate(Result, into = c("PlayerScore", "OpponentScore"), sep = "-", convert = TRUE) %>%
  mutate(
    Outcome = case_when(
      PlayerScore > OpponentScore ~ "Win",
      PlayerScore < OpponentScore ~ "Loss",
      PlayerScore == OpponentScore ~ "Draw"
    ),
    Win = ifelse(PlayerScore > OpponentScore, 1, 0)
  )
CODGames2_mp_HC_TDM

full_model <- lm(Score ~ TotalXP + Eliminations + Deaths + Damage + XPType + Win, data = CODGames2_mp_HC_TDM)

summary(full_model)
```

NOTE: The variable Damage seems to be insignificant as it has a large p-value (> 0.05)

## Task 2a. LASSO Regresison and Other Feature Selection Method

```{r}
Xmat <- model.matrix(Score ~ TotalXP + Eliminations + Deaths + Damage + XPType + Win, data = CODGames2_mp_HC_TDM)[ , -1]
Ymat <- CODGames2_mp_HC_TDM$Score

lasso_model <- glmnet(Xmat, Ymat, alpha = 1)

set.seed(123)
lassoCV <- cv.glmnet(Xmat, Ymat, alpha = 1)
set.seed(NULL)

plot(lassoCV)
best_lambda <- lassoCV$lambda.min
best_lambda

coef(lassoCV, s = "lambda.min")
```

LASSO regression selected a value of $\lambda = 2.611974$ which produced the lowest MSE value. This produced the equation: `Score` = 937.15693473 + 0.05984054`TotalXP` + 159.12937918`Eliminations` - 72.77820466`Deaths` + 0.94772672`Damage` - 361.57550500`XPTypeDouble XP + 10%` - 447.19199686`Win`.

```{r}
stepwise_model <- stepAIC(full_model, direction = "both")
summary(stepwise_model)
```

Using step-wise selection, it produced the equation: `Score` = 944.05302 + 0.06054`TotalXP` + 185.24081`Eliminations` - 73.25104`Deaths` - 367.94000`XPTypeDouble XP + 10%` - 454.44701`Win`. We can see that the step-wise model removed the variable `Damage` while the LASSO regression did not. The coefficients of the variables for both models have the same sign and seem to be fairly close in value.

## Task 2b. Regression Tree

```{r}
RegTree <- rpart(Score ~ TotalXP + Eliminations + Deaths + Damage + XPType + Win,
                 method = "anova",
                 data = CODGames2_mp_HC_TDM,
                 minbucket = 3,
                 control = rpart.control(minbucket = 15)
)

summary(RegTree)

fancyRpartPlot(RegTree, cex = 1)

RegTree$variable.importance
```

Based on the regression tree and using RegTree$variable.importance, we were able to get the values associated with each variable's importance with the three highest variable importance values being `Damage` (250835300), `Eliminations` (237532841), and `TotalXP` (64376422).

## Task 2c. Standardized Model

```{r}
CODGames2_mp_HC_TDM_standardized <- CODGames2_mp_HC_TDM %>%
  mutate(XPType = ifelse(XPType == "Double XP + 10%", 1, 0),
         TotalXP = scale(TotalXP, center = TRUE, scale = TRUE),
         Eliminations = scale(Eliminations, center = TRUE, scale = TRUE),
         Deaths = scale(Deaths, center = TRUE, scale = TRUE),
         XPType = scale(XPType, center = TRUE, scale = TRUE),
         Win = scale(Win, center = TRUE, scale = TRUE)
         )

standardized_model <- lm(Score ~ TotalXP + Eliminations + Deaths + XPType + Win, data = CODGames2_mp_HC_TDM_standardized)
summary(standardized_model)
```

Using the standardized model, it produced the estimated equation: `Score` = 3104.95 + 382.88`TotalXP` + 1030.92`Eliminations` - 276.49`Deaths` - 182.80`XPType` - 226.12`Win`. Based on the magnitude (absolute value) of the estimated coefficients, the three most important variables are `Eliminations` (1030.92), `TotalXP` (382.88), and `Deaths` (276.49).

# Contribution
Jackie Yang: Completed task1 and task2a and task2b and their comment

Xinyu Xie: Completed task2c with comments and fixing some problems
